---
title: "Sok-1005_Assignment_1"
author: Daniel Johannessen. kandidat nummer 3
format: html
editor: visual
---

## Assignment 1 Reading static webpages.

We start by gathering the data we need and getting all our packages

The data is: Lower-Troposphere\
Mid-Troposphere\
Tropopause\
Lower-Stratosphere

```{r, message=False, warning=False}
#above this line i specify that i do not want quarto to output the message and the warning
rm(list=ls()) #emptying enviroment
library(tidyverse) #We load in packages
library(lubridate)
library(zoo)


lower <- read_table("http://vortex.nsstc.uah.edu/data/msu/v6.0/tlt/uahncdc_lt_6.0.txt") #We collect our data and save it into datasets
mid <- read_table("http://vortex.nsstc.uah.edu/data/msu/v6.0/tmt/uahncdc_mt_6.0.txt")
tropopause <- read_table("http://vortex.nsstc.uah.edu/data/msu/v6.0/ttp/uahncdc_tp_6.0.txt")
stratosphere <- read_table("http://vortex.nsstc.uah.edu/data/msu/v6.0/tls/uahncdc_ls_6.0.txt")
```

Now that we have gotten our data and our packages, i start with removing the last rows in the dataset that contains comments that we don't need.

Here I use code to dynamically remove the comments even if new data is added. The code is gathered from the link provided below. It is at line 43.

[Source of the code used for dynamic code](https://github.com/uit-sok-1005-v23/uit-sok-1005-v23.github.io/blob/main/Global_Temperature.R "Line 43")

```{r}

lower <- lower[1:which(lower$Year %in% "Year")-1, ]
mid <- mid[1:which(mid$Year %in% "Year")-1, ]
tropopause <- tropopause[1:which(tropopause$Year %in% "Year")-1, ]
stratosphere <- stratosphere[1:which(stratosphere$Year %in% "Year")-1, ] #We remove all rows below the second "Year" in the table.


```

We are told to also only use data in the Globe column so I will now select it out of each dataset then combine the datasets into one.

```{r}
lower <- lower %>% 
  mutate(Date = ym(paste(Year, Mo))) %>% #Creating a Date variable
  select(Date , Year, Mo, Globe) %>%  #selecting and arranging the data we want
  rename(lower_globe = Globe) #At the end we rename our variables
  
mid <- mid %>% 
  mutate(Date = ym(paste(Year, Mo))) %>% 
  select(Date ,Year, Mo, Globe) %>% 
  rename(mid_globe = Globe)

stratosphere <- stratosphere %>% 
  mutate(Date = ym(paste(Year, Mo))) %>% 
  select(Date, Year, Mo, Globe) %>% 
  rename(stratosphere_globe = Globe)

tropopause <- tropopause %>% 
  mutate(Date = ym(paste(Year, Mo))) %>% 
  select(Date, Year, Mo, Globe) %>% 
  rename(tropopause_globe = Globe)

```

We now combine them into one dataset we can work with and we convert strings to numeric.

We also create new variables for our averages per year

```{r}

df <- lower %>% 
  cbind(mid_globe = mid$mid_globe) %>% #Cbind is binding the columns from the other dataset into this dataset we work with
  cbind(stratosphere_globe = stratosphere$stratosphere_globe) %>% 
  cbind(tropopause_globe = tropopause$tropopause_globe) %>% 
  mutate_if(is.character, ~as.numeric(.)) #finally we convert characters to numeric




```

Now that we have made our datasets, its time to start creating rolling averages

```{r}

rolling <- df %>% 
  mutate(rolling_lower = rollmean(lower_globe, 12,  fill=NA, align="right")) %>% #I creating a new variable for the rolling average. 
  mutate(rolling_mid = rollmean(mid_globe, 12, fill=NA, align="right")) %>% 
  mutate(rolling_strato = rollmean(stratosphere_globe, 12, fill=NA, align="right")) %>% 
  mutate(rolling_tropo = rollmean(tropopause_globe, 12, fill=NA, align="right")) 


```

```{r}

rolling <- rolling %>% 
  mutate(average = (rolling_lower + rolling_mid + rolling_strato + rolling_tropo)/4) #Here i create a new variable for the average by combining all 4 datasets and dividing by 4



```

Now that we have our dataset with our rolling averages and our average of the rolling averages its time to start plotting.

```{r, warning=False}
rolling %>% 
  filter(Year >= 1980) %>% #filtering for years from and including 1980
  ggplot(aes(x=Date))+ #we set our x axis to date
  geom_line(aes(y=rolling_lower, color="Lower Troposphere"))+ #we set lines for all the different variables
  geom_line(aes(y=rolling_mid, color="Mid-Troposphere"))+
  geom_line(aes(y=rolling_strato, color="Lower Stratosphere"))+
  geom_line(aes(y=rolling_tropo, color="Tropopause")) +
  geom_line(aes(y=average, color="Centered average"), size=1.2)+
  labs(title="Global temperature changes over time",#Setting title
  color="Location", #Changing legend name to location
  y="Temperature in celsius", #Setting Y axis name
  x= "Year",  #Setting X axis name
  subtitle="The graph shows the temperature deviation from the 1991-2020 average.
The graph shows right aligned rolling averages", #Setting subtitle
  caption="Source: www.drroyspencer.com/latest-global-temperatures/")+ #Setting caption
  theme_bw() + #we set the theme to black and white instead of default 
  theme(plot.caption.position = "plot",
        plot.caption = element_text(hjust = 1))#Setting caption location
```
